---
title: "EDO prédation"
format: html
editor: visual
---

```{r}
## Packages
library(ggplot2)
library(dplyr)
library(tidyr)
```

**INTRODUCTION** :

When having a dynamic system modeled by an ODE, we usually can't find a mathematical solution. Thus, we have 2 options : a numerical approach and a qualitative one. Here, we will focus on the latter, that is to say qualitatively predict the evolution (in terms of behavior) of N(t) (the number of the population at time t), in particular asymptotically.

Looking at lynx and hare pelts from the Hudson's Bay Company's (fur-collecting) trappers' hunts over a 90-year period, we see that the number of hare and lynx pelts is correlated. If we assume that the number of pelts collected is proportional to the total population, then we arrive at the conclusion that the hare and lynx population is correlated with a phase shift of a few years for the lynx.

What is the intuition behind this? Lynxes need hares to survive, and so the lynx population can only grow if the hare population grows. There's a time lag: when the lynx population is low, the hare population begins to grow. The lynxes then begin to have food again, and will then grow again and begin to massacre a large part of the hare population, only to run out of food. Then the cycle begins again.

**Logistic Prey-Predator model (in a limited environment)** :

$$
\begin{cases} 
\frac{dN}{dt}=r_1N(1-\frac{N}{K})-b_1NP \\
\frac{dP}{dt}=-r_2P+b_2NP
\end{cases}
$$

The parameters are the following ones (they are all positive :) ) :

| Parameters | Meaning                             |
|------------|-------------------------------------|
| $r_{1}$    | growth rate of preys' population    |
| $r_{2}$    | death rate of predators' population |
| $b_{1}$    | predation efficiency                |
| $b_{2}$    | predation yield                     |
| ${K}$      | biotic capacity                     |

In such conditions, we expect the populations dynamics to evolve in different ways according to parameters value and initials conditions (the size of both population). We can relatively easily assume that a big predator's population paired with a high predation efficiency would lead prey's population to go extinct and, shortly after, predator's population is expected to follow the same dynamic. The question we may now ask ourselves is the following : how would the system evolve according to parameters variation? This is where qualitative analysis steps in. But first let's do a little bit of math.

**Equilibrium points and isoclines**

Let's start with a bit of vocabulary, in qualitative analysis we call equilibrium points all value of N for which the derivative is 0. Mathematicaly it's expressed as following : $$
\frac{dN}{dt}=0\;\Longleftrightarrow\;f(N^{*})=0 \\
\text{with}\;\ \frac{dN}{dt}=f(N)
$$ For equation system such as Lokta-Voltera it would be as following : $$
\begin{cases} \frac{dN_{1}}{dt}=0 \\ \frac{dN_{2}}{dt}=0 \end{cases} \;\Longleftrightarrow\; \begin{cases} f_{1}(N1^{*},N2^{*})=0 \\ f_{2}(N1^{*},N2^{*})=0 \end{cases}\\
\text{with}\;\ \begin{cases} \frac{dN_{1}}{dt}=f_{1}(N1,N2) \\ \frac{dN_{2}}{dt}=f_{2}(N1,N2) \end{cases}
$$ Those equilibrium points carry the information about the population sizes at which the system does not change. They can be either stable (LAS) unstable (repeller) and in the case the system has more than one differential equation, saddle points may appear. Around LAS points, system will go towards the equilibrium, around repeller points, system will tend to move away from equilibrium. As for saddle points, depending on the area the system will go toward or away the equilibrium points.

In contrast to equilibrium points, where all derivative must be equal to zero, isoclines correspond to lines for which an equation of the system is individually equal to zero. Mathematicaly it's expressed as following :

$$
\text{Isoclines N1 :} \\
\frac{dN_{1}}{dt}=0 \;\Longleftrightarrow\; f_{1}(N1,N2)=0 \\
\text{Isoclines N2 :}\\
\frac{dN_{2}}{dt}=0\;\Longleftrightarrow\; f_{2}(N1,N2)=0 \\
\text{with}\;\begin{cases} \frac{dN_{1}}{dt}=f_{1}(N1,N2) \\ \frac{dN_{2}}{dt}=f_{2}(N1,N2) \end{cases}
$$

Plotting those isoclines on a phase plane allows us to identify some areas where the system may evolve in different ways. We can then examine the behavior of each population in the previously defined area , allowing to characterize the system overall behavior.

All of that is analytically calculable for some systems of differential equation system (including Lotka-Volterra logistic), but lazy as we are, we will focus on numerically approximate those solutions.

**Numerical approximation of equilibrium points and phase plane**

```{r}
Proiepred<- function(t, y, parameters) {
  N <- y[1]
  P <- y[2]
  r1 <- parameters[1]
  r2 <- parameters[2]
  b1 <- parameters[3]
  b2 <- parameters[4]
  dPop <- numeric(2)
  dPop[1] <- r1*N - b1*N*P
  dPop[2] <- - r2*P + b2*N*P
  list(dPop)
}
```

```{r}
##Fonction simulation trajectoires
equation_proie <- function(N,P,r1,b1,K){
  return(r1*N*(1-(N/K))-b1*N*P)
}

equation_predateur <- function(N,P,r2,b2){
  return(-r2*P+b2*N*P)
}

## N_0   : effectif des proies au temps 0 
## P_0   : effectif des prédateurs au temps 0
## t_end : temps final de la simulation
## pas   : pas de temps
## r1    : taux de reproduction intrinsèque des proies
## b1    : taux de mortalités des proies dû aux prédateurs
## r2    : taux de reproduction des prédateur en fonction des proies 
## b2    : taux de mortalité intrinsèques des prédateurs
## K     : capacité de charge des proies
simul_traj <- function(N_0,P_0,t_end,pas,r1,b1,r2,b2,K){
  list_N <- c(N_0)
  dN <- N_0
  list_P <- c(P_0)
  dP <- P_0
  list_t <- c(0)
  t <- 0
  while (t<t_end){
    t<- t+pas
    list_t <- c(list_t,t)
    dN <- equation_proie(dN,dP,r1,b1,K)*pas
    if (dN<0){dN <- 0}
    dP <- equation_predateur(dN,dP,r2,b2)*pas
    if (dP<0){dP <- 0}
    list_N <- c(list_N,dN)
    list_P <- c(list_P,dP)
  }
  #return(list(list_N=list_N,list_P=list_P,list_t=list_t))
  
  ##data_wrangling
  data <- data.frame(time=list_t,prey=list_N,predator=list_P)
  data <- gather(data,key="specie",value="number",prey:predator)
  
  ##plot
  plot <- ggplot(data,aes(x=time,y=number,color=specie,group=specie)) +
    geom_line()
  plot
}

##Just for the example of the function
simul_traj(100,5,100,1,1.5,0.05,0.5,0.05,500)


```
