---
title: "EDO Estimation of parameters : the logistic growth"
format: html
editor: visual
---

# Introduction


This document aims to be a presentation of a logistic growth simulation: from the data simulation to the estimation of the parameters.

The following document will thus be organized in two parts : the data simulation and the estimation of the parameters.

# Data Simulation

```{r}
# Parameters
time <- seq(0, 25, 0.25)
r <- 0.5
K <- 100
error <- 5
y0 <- 1

# Logistic growth function with error
logistic <- function(t, r, K, y0, error) {
  growth <- K / (1 + ((K - y0) / y0) * exp(-r * t))
  return(growth + rnorm(1, mean = 0, sd = error))  # Add random error
}

# Apply the logistic function over the time vector
y <- sapply(time, logistic, r = r, K = K, y0 = y0, error = error)

# Plot the result
plot(time, y, col = "blue", xlab = "Time", ylab = "Population", main = "Logistic Growth with Error")
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

# Estimation of the parameters

## General theory
